Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-26 paper/perfmod.tex
edits

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit Stokes geometry/velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$n$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

We parameterize simulation performance using either $\Delta x$, the representative horizontal mesh/grid cell diameter, i.e.~the resolution, or $n$, the number of nodes or vertices in the horizontal mesh or grid.  The model domain is of (horizontal) width $L$ so $\Delta x = O(L n^{-1/2})$; there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$, but our analysis focusses on spatially-3D models with map-plane horizontal meshes or grids.}  Table \ref{tab:notation} lists the parameters used in our analysis.  Big-O notation applies only in the high resolution limit where $\Delta x \to 0$ and $n\to \infty$.

A standard numerical ice sheet model uses $n$ ice thickness or surface elevation variables, the number of \emph{degrees of freedom}, and it requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This bounded-vertical-nodes assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.  However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance can recompute velocity as a function of (true) state variables.

>>>>

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$n$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

Table \ref{tab:notation} lists the parameters used in our simulation performance  analysis.  Our primary parameters are $\Delta x$, a representative value for the horizontal mesh (grid) cell diameter, and $n$, the number of nodes (vertices) in the horizontal mesh.  \emph{High resolution} refers to the equivalent $\Delta x\to 0$ and $n\to\infty$ limits, and big-O notation is only used in this limit.  If the model domain is of width $L$ then these parameters are related by
\begin{equation}
\Delta x = O\left(\frac{L}{\sqrt{n}}\right) \quad \leftrightarrow \quad n = O\left(\frac{L^2}{\Delta x^2}\right).  \label{eq:paramrelation}
\end{equation}
That is, for a given value of $L$ there are $O(\sqrt{n})$ mesh cells in each horizontal dimension.\footnote{In flow-line models $\Delta x = O(L n^{-1})$.  However, our analysis addresses spatially-3D models with map-plane horizontal meshes.}

A numerical ice sheet model uses $n$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(n)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.  However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(n)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(n)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.
